---
title: "08_Mm_responsive_genes"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
#library(knitr)
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
#library(d3heatmap)
#library(flexdashboard)
library(edgeR);library(tidyverse)
library(readr);library(readxl)
library(stringr)

```

# mapping read files by Kalisto

# working in directories containing fastq files (working in Whitney/NGS/Arabidopsis_analysis_3/Ro)
```{r eval=FALSE}
fastqfiles<-list.files(pattern="fastq")
system(paste("fastqc ",fastqfiles,sep=""))
```

Build index
```{r eval=FALSE}
system("kallisto index -i TAIR10_cdna_20110103_representative_gene_model_kallisto_index TAIR10_cdna_20110103_representative_gene_model.gz")
# This reference seq has been updated in 2012! Redo mapping (101018)
system("kallisto index -i TAIR10_cdna_20110103_representative_gene_model_updated_kallisto_index TAIR10_cdna_20110103_representative_gene_model_updated")
```
mapping
```{r eval=FALSE}
# working in fastq file directory
# making directory for mapped 
system("mkdir ../kallisto_sam_out2")
# go to fastq file directory
# reading fastq files (in fastq file directory)
#fastqfiles<-list.files(pattern="fastq.gz")
# fastq files with single end 
# fastqfiles.SE<- fastqfiles %>% as_tibble() %>% separate(value,into=c("SRA","type","compress"),sep="\\.") %>% separate(SRA,into=c("SRA","pair"),sep="_") %>% group_by(SRA) %>%summarize(num=n()) %>% filter(num==1) %>% select(SRA) %>% as_vector()
for(x in fastqfiles) {
system(paste("kallisto quant -i ../references/TAIR10_cdna_20110103_representative_gene_model_updated_kallisto_index  -o ../kallisto_sam_out2/",x," --single -l 250 -s 40 --pseudobam ",x," | samtools view -Sb - > ../kallisto_sam_out2/",x,"/",x,".bam",sep=""))
setwd(paste("../kallisto_sam_out2/",x,sep=""))
system(paste("samtools sort *.bam -o ",x,".sorted.bam",sep=""))
system("samtools index *.sorted.bam")
setwd("../../Rodriguez-Celma_2016_RNAseq")
}
```

Normalize reads and create DGE objects

```{r}
library(tidyverse)
library(edgeR)
```

# copy kallisto_sam_out2 directory in Whitney under input directory
## get reads and create data frame

```{r}
files <- dir("../input/kallisto_sam_out2/", include.dirs = TRUE)
files %>% head()
```

```{r}
counts.list <- map(files, ~ read_tsv(
  file=file.path("..","input","kallisto_sam_out2",.,"abundance.tsv"),
  col_types = "cdddd"))
names(counts.list) <- files
```

```{r}
counts.RodriguezCelma2016 <- sapply(counts.list, dplyr::select, est_counts) %>% 
  bind_cols(counts.list[[1]][,"target_id"],.)
counts.RodriguezCelma2016[is.na(counts.RodriguezCelma2016)] <- 0
colnames(counts.RodriguezCelma2016) <- sub(".fastq.gz.est_counts","",colnames(counts.RodriguezCelma2016),fixed = TRUE)
counts.RodriguezCelma2016
```

```{r}
write_csv(counts.RodriguezCelma2016,"../output/2016-Rodriguez-Celma_2016_raw_counts.csv.gz")
```

```{r}
counts2 <- counts.RodriguezCelma2016 %>% 
  as.data.frame() %>% 
  column_to_rownames(var = "target_id") %>%
  as.matrix() %>%
  round(0)
```
# dge (under construction)
```{r}
dge <- DGEList(counts=counts2, 
               samples=colnames(counts2), 
               group=factor(rep(c("minusMn","ES"),each=3),levels=c("ES","minusMn"))
                                )
```

# normalize

```{r}
dge <- calcNormFactors(dge)
```
# save
```{r}
save(dge, file="../output/Rodriguez-Celma_2016.Rdata")
```
# DEG
```{r}
group<-factor(rep(c("minusMn","ES"),each=3),levels=c("ES","minusMn"))
design <- model.matrix(~group)
y <- estimateDisp(dge,design)
fit <- glmFit(y,design)
lrt <- glmLRT(fit,coef=2)
Mn_deficiency.DEGs.all<-topTags(lrt,n = Inf,p.value = 1)$table

```
# Adding annotation 
```{r}
#At.gene.name <-read_tsv("https://www.arabidopsis.org/download_files/Subscriber_Data_Releases/TAIR_Data_20180630/gene_aliases_20180702.txt.gz") # Does work from home when I use Pulse Secure.
# At.gene.name <- 
#read_tsv("https://www.arabidopsis.org/download_files/Public_Data_Releases/TAIR_Data_20181231/Araport11_functional_descriptions_20181231.txt.gz") # does not exist anymore
# updated one (Jan 21, 2020)
#At.gene.name<-read_tsv("https://www.arabidopsis.org/download_files/Subscriber_Data_Releases/TAIR_Data_20191231/gene_aliases_20191231.txt.gz") # download file and remove all "" with TextWrangler etc
At.gene.name<-read_tsv(file.path("..","Annotation_copy","input","gene_aliases_20191231.txt")) 
# combine redundant symbol
At.gene.name <- At.gene.name %>% group_by(name) %>% summarise(symbol2=paste(symbol,collapse=";"),full_name=paste(full_name,collapse=";"))
At.gene.name %>% dplyr::slice(100:110)
At.gene.name %>% filter(name=="AT4G25350") # SHB1
# add annotiona with full name objects (all)
Mn_deficiency.DEGs.all.anno <- Mn_deficiency.DEGs.all %>% rownames_to_column(var="genes") %>% separate(genes,into=c("AGI","variant"),sep="\\.", remove=FALSE) %>% left_join(At.gene.name,by=c(AGI="name")) %>%  dplyr::select(AGI, logFC,logCPM,LR,PValue,FDR,symbol2, full_name)
Mn_deficiency.DEGs.all.anno %>% dplyr::slice(1:100) # %>% View()
# write csv (all)
write.csv(Mn_deficiency.DEGs.all.anno,file=file.path("..","output","Mn_deficiency.DEGs.all.anno.csv"))
```

# Expression pattern
```{r}
load("../output/Rodriguez-Celma_2016.Rdata") # this is wrong

Mn.RodriguezCelma_2016.cpm<-cpm(dge) %>% as_tibble() %>% bind_cols(data.frame(transcript_ID=rownames(dge$counts)),.) %>% separate(transcript_ID,into=c("AGI","variant"),remove=FALSE)
sample.description.Mn.RodriguezCelma.2016<-dge$samples %>% as_tibble() %>% dplyr::rename(sample=samples)

expression.pattern.AT.graph.cpm<-function(data=Mn.RodriguezCelma_2016.cpm,target.genes.FDR,sample.description=sample.description.Mn.RodriguezCelma.2016,title=""){
  data[is.na(data)] <- 0 #
  # using FDR
  data.temp<-data  %>% 
    dplyr::rename(target_id=AGI) %>%
    filter(target_id %in% target.genes.FDR$AGI) %>% dplyr::select(-transcript_ID,-variant) %>%
    gather(sample,value,-target_id) %>%
    inner_join(sample.description, by="sample")  %>%
    right_join(target.genes.FDR[,c("AGI","FDR")], by=c("target_id"="AGI")) %>%
    mutate(FDR=format(FDR,digits=2,scientific=TRUE)) %>%
    unite(target_id.FDR,target_id,FDR,sep="\n FDR ") 
  # print(data.temp)
    p<-data.temp %>% ggplot(aes(x=group,y=value))  + 
      geom_jitter(alpha = 0.5,aes(color=group))  + 
      theme_bw() + facet_grid(target_id.FDR~.,scales="free") +
      theme(strip.text.y=element_text(angle=0),axis.text.x=element_text(angle=90)) +
      theme(legend.position="bottom") + labs(title="")
    p
}
# expression pattern of upregulated genes upon Mn deficienty
   gene.of.interest.FDR.up <- Mn_deficiency.DEGs.all.anno %>% filter(FDR< 0.05,logFC>0) %>%
     dplyr::select(AGI,logFC,FDR) %>% arrange(FDR)
   print("logFC>0")
expression.pattern.AT.graph.cpm(target.genes.FDR=gene.of.interest.FDR.up[1:5,])
# down-regulated genes upon Mn deficienty
   print("logFC<0")
gene.of.interest.FDR.down <- Mn_deficiency.DEGs.all.anno %>% filter(FDR< 0.05,logFC<0) %>%
     dplyr::select(AGI,logFC,FDR) %>% arrange(FDR)
expression.pattern.AT.graph.cpm(target.genes.FDR=gene.of.interest.FDR.down[1:5,])
```


# Session info
```{r}
sessionInfo()
```
