---
title: "09_clustering"
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: inline
---

Original script by Kazu, updated by Julin.  This version if for WGCNA only on the samples treated with microbes.

```{r setup}
getwd()
knitr::opts_chunk$set(echo = TRUE,error=TRUE)
library(tidyverse)
library(Biostrings)
library(goseq)
library(GO.db)
library(patchwork)
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")

# read functions for expression graph
#source("../tools_copy/Expression_pattern_graph.R")
```

get data

```{r}
load("../output/timecourseDGE.Rdata")
cpm.timecourse.v3.0 <- cpm(dge) %>% as_tibble() %>% bind_cols(data.frame(transcript_ID=rownames(dge$counts)),.)
cpm.timecourse.v3.0.log <- cpm(dge,log=TRUE) %>% as_tibble() %>% bind_cols(data.frame(transcript_ID=rownames(dge$counts)),.)

sample.description.timecourse <- dge$samples %>% bind_cols(data.frame(sample=rownames(dge$samples)),.)
```

Filter to keep only live samples
```{r}
sample.description.timecourse.live <- sample.description.timecourse %>% filter(soil_trt=="SBC_OLD")
sample.description.timecourse.live

cpm.timecourse.v3.0.live <- cpm.timecourse.v3.0 %>% select(transcript_ID, all_of(sample.description.timecourse.live$sample))

cpm.timecourse.v3.0.log.live <- cpm.timecourse.v3.0.log %>% select(transcript_ID, all_of(sample.description.timecourse.live$sample))

```



```{r}
# filtering for selecting only expressed genes
cpm.timecourse.v3.0.live <- cpm.timecourse.v3.0.live[rowSums(cpm.timecourse.v3.0.live[,-1]>5) >=14,] # 10% of libraries have
dim(cpm.timecourse.v3.0.live) # [1] 21017   145

# removing low expressed genes from cpm.timecourse.v3.0.log.live
cpm.timecourse.v3.0.log.live <- cpm.timecourse.v3.0.log.live %>% filter(transcript_ID %in% cpm.timecourse.v3.0.live$transcript_ID)
dim(cpm.timecourse.v3.0.log.live) #[1] 21017   289
```

# format data

# use log transformed data
```{r}
co.var.df <- function(x) ( 100*apply(x,1,sd)/rowMeans(x) )

cpm.timecourse.v3.0.log.live$cv <- co.var.df(cpm.timecourse.v3.0.log.live[,-1])

cpm.timecourse.v3.0.log.live %>% arrange(desc(cv))

b <- hist(log10(cpm.timecourse.v3.0.log.live$cv))
b

# use largeCV 
cpm.timecourse.v3.0.log.live.largeCV <- cpm.timecourse.v3.0.log.live[abs(cpm.timecourse.v3.0.log.live$cv) > 10,] 

dim(cpm.timecourse.v3.0.log.live.largeCV) # [1] 9499  146

hist(log10(cpm.timecourse.v3.0.log.live.largeCV$cv))

cpm.timecourse.v3.0.log.live.largeCV <- cpm.timecourse.v3.0.log.live.largeCV %>%
  select(-cv)
###########

write_csv(cpm.timecourse.v3.0.log.live.largeCV,file=file.path("..","output","cpm.timecourse.v3.0.log.live.largeCV.csv.gz"))
```

# WGCNA
## co-expression analysis by WGCNA
```{r eval=TRUE}
# The following setting is important, do not omit.
library(WGCNA) 
options(stringsAsFactors = FALSE)
```

```{r eval=TRUE}
cpm.timecourse.v3.0.log.live.largeCV <- read_csv(file.path("..","output","cpm.timecourse.v3.0.log.live.largeCV.csv.gz"))

# 
datExpr <-t(cpm.timecourse.v3.0.log.live.largeCV[,-1])

# Choose a set of soft-thresholding powers
powers = c(c(1:9), seq(from = 10, to=20, by=2))
sft = pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)  

# Plot the results:
#sizeGrWindow(9, 5)
par(mfrow = c(1,2));
cex1 = 0.9;
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")

# make modules 
conflicted::conflicts_prefer(WGCNA::cor)
net = blockwiseModules(datExpr, power = 5,
                       maxBlockSize = 10000,
                       networkType = "signed hybrid",
                       TOMType = "unsigned", minModuleSize = 20,
                       reassignThreshold = 0, mergeCutHeight = 0.25,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       saveTOMs = TRUE,
                       saveTOMFileBase = "cpm.timecourse.v3.0.log.live.largeCV.TOM",
                       verbose = 3)
save(net,file="../output/WGCNA/net.cpm.timecourse.v3.0.log.live.largeCV.Rdata")  
# open a graphics window
pdf(file="../output/WGCNA/live_largeCV.dendrogram.pdf",width=10,height=8)
# Convert labels to colors for plotting
mergedColors = labels2colors(net$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(net$dendrograms[[1]], mergedColors[net$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
dev.off()
# save parameters  
moduleLabels = net$colors
moduleColors = labels2colors(net$colors)
MEs = net$MEs
colnames(MEs) <- colnames(MEs) %>% str_remove("ME") %>% as.numeric() %>% labels2colors()
geneTree = net$dendrograms[[1]]
save(sample.description.timecourse.live, MEs, moduleLabels, moduleColors, geneTree,file ="../output/WGCNA/live.largeCV.RData")

```

```{r, eval=FALSE}
load("../output/WGCNA/live.largeCV.RData")
```


```{r}

# how many modules?
table(net$colors)

length(table(net$colors)) # 26 modules

```

  
  
# adding gene name, annotations
```{r}
cpm.timecourse.v3.0.log.live.largeCV.modules <- tibble(
  transcript_ID=cpm.timecourse.v3.0.log.live.largeCV$transcript_ID,
  module=moduleColors
)


## prep
# annotation file for v3.0annotation

Br.v3.0.At.BLAST <- read_csv(file.path("..","Annotation_copy","output","v3.0annotation","Brapa_v3.0_annotated.csv")) 

# This annotation is redundant with name (Br grene). Eg 
Br.v3.0.At.BLAST %>% filter(name=="BraA01g040570.3C")
# reduce the redundancy (112418)
Br.v3.0anno.At.BLAST.highscore <- Br.v3.0.At.BLAST %>% group_by(name) %>% slice_max(score)

cpm.timecourse.v3.0.log.live.largeCV.modules.annotated <- cpm.timecourse.v3.0.log.live.largeCV.modules %>%
  left_join(Br.v3.0anno.At.BLAST.highscore, by=c("transcript_ID" = "name")) %>%
  select(AGI, At_symbol, At_full_name, At_short_description, At_Curator_summary, At_Computational_description, perc_ID, aln_length)

```


Gene lengths
```{r, eval=FALSE}
library(Biostrings)
cDNAs <- readDNAStringSet("~/Sequences/ref_genomes/B_rapa/genome/V3.0/Brapa_genome_v3.0_cds.fasta.gz")
CDS.lengths <- nchar(cDNAs)
names(CDS.lengths) <- names(cDNAs)
head(CDS.lengths)
save(CDS.lengths, file="../Annotation_copy/output/v3.0annotation/Brapa_V3.0_CDS_lengths.Rdata")
```

Load lengths and annotations
```{r}
load("../Annotation_copy/output/v3.0annotation/Brgo.v3.0anno.Atgoslim.BP.list.Rdata") # object name `Brgo.v3.0anno.Atgoslim.BP.list`
load("../Annotation_copy/output/v3.0annotation/Brapa_V3.0_CDS_lengths.Rdata") # object name `CDS.lengths`
```

Set up gene list
```{r}
all.expressed.genes.modules <- cpm.timecourse.v3.0.log.live %>% 
  select(transcript_ID) %>%
  full_join(cpm.timecourse.v3.0.log.live.largeCV.modules) %>%
  mutate(module=ifelse(is.na(module), "none", module))

all.expressed.genes.modules

CDS.lengths <- CDS.lengths[all.expressed.genes.modules$transcript_ID]

length(CDS.lengths) == nrow(all.expressed.genes.modules)
```

save some results

```{r}
all.expressed.genes.modules.ananotated <- all.expressed.genes.modules %>% 
  left_join(Br.v3.0anno.At.BLAST.highscore, by=c("transcript_ID" = "name")) %>%
  select(AGI, At_symbol, At_full_name, At_short_description, At_Curator_summary, At_Computational_description, perc_ID, aln_length)

write_csv(all.expressed.genes.modules.ananotated, file = "../output/WGCNA/live.expressed.genes.modules.annotated_all.genes.csv")
write_csv(all.expressed.genes.modules, file = "../output/WGCNA/live.expressed.genes.modules_all.genes.csv")

```


```{r, message=FALSE}
modules.to.gene.vector <- function(module, module.df=all.expressed.genes.modules) {
  module.vector <- as.numeric(module.df$module==module)
  names(module.vector) <- module.df$transcript_ID
  module.vector
}

# Calculate GO enrichment
go.results <- tibble(module=sort(unique(all.expressed.genes.modules$module))) %>%
  mutate(module.gene.vector = map(module, modules.to.gene.vector),
         module.nullp = map(module.gene.vector, ~ nullp(.x, bias.data=CDS.lengths, plot.fit=FALSE)),
         module.goseq = map(module.nullp, ~goseq(.x, 
                                                 gene2cat = Brgo.v3.0anno.Atgoslim.BP.list , 
                                                 test.cats = "GO:BP"))
  )

# Calculate FDR
go.results <- go.results %>%
  mutate(module.goseq = map(module.goseq,
                            ~ mutate(.x, over_represented_FDR = p.adjust(over_represented_pvalue,
                                                                         method = "fdr"))))

# Filter based on FDR
go.results <- go.results %>%
  mutate(module.goseq = map(module.goseq,
                            ~ filter(.x, over_represented_FDR <= 0.05)))

# unnest
go.results <- go.results %>% dplyr::select(module, module.goseq) %>%
  unnest(module.goseq)

write_csv(go.results, "../output/WGCNA/GO_ORA_WGCNA_live_genes.csv")

go.results

go.top6 <- go.results %>% 
  group_by(module) %>%
  slice_min(over_represented_FDR, n=6, with_ties = FALSE) %>%
  summarize(top_6_go = str_c(term, sep=", ", collapse = ", ")) %>%
  mutate(top_6_go = str_replace(top_6_go, "(^(.+?, ){3})","\\1\n"))

go.top6
```



# Plotting modules

set up data frame
```{r}
MEs.plot <- MEs %>% rownames_to_column("sample") %>%
  full_join(sample.description.timecourse.live, .) %>%
  select(-lib.size, -norm.factors, -block, -group, -group.1)

MEs.plot
```

## Find modules reponsive to day or time 

```{r}
MEs.nest <- MEs.plot %>%
  select(-genotype, -sample) %>%
  pivot_longer(cols = -(soil_trt:sampling_time), names_to = "module", values_to = "ME_expression") %>%
  arrange(module, sampling_day, sampling_time, soil_trt) %>%
  nest(.by = module)

MEs.nest
```

```{r}
MEs.trt.aov <- MEs.nest %>%
  mutate(aov = map(data, ~ aov(ME_expression ~ sampling_day*sampling_time, data = .x)),
         terms = map(aov, broom::tidy)) %>%
  select(module, terms) %>%
  unnest(terms) %>%
  filter(term != "Residuals")  %>%
  mutate(term=str_remove_all(term, "_trt|sampling_")) # clean up the names

# any terms not significant?

MEs.trt.aov %>%
  summarize(min_p = min(p.value), .by = term)

# All Sig


# Caclulate FDR

MEs.trt.aov <- MEs.trt.aov %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value, method = "fdr")) %>%
  ungroup()

MEs.trt.aov %>%
  summarize(min_fdr = min(fdr), .by = term) 

```

most significant day responders
```{r}
MEs.trt.aov %>%
  filter(term=="day", fdr < 0.05) %>%
  arrange(fdr)
```

most signficant time responders
```{r}
MEs.trt.aov %>%
  filter(term=="time", fdr < 0.05) %>%
  arrange(fdr)
```

## plot one set of fdrs

```{r}
MEs.trt.aov %>% filter(module=="blue")
```


```{r}
pl1 <- MEs.trt.aov %>% filter(module=="blue") %>%
  ggplot(aes(y=fdr, x = term)) +
  geom_col(fill="blue") + 
  scale_y_continuous(transform=c("log10", "reverse")) +
  scale_x_discrete(limits = c("day", "time", "day:time")) +
  geom_hline(yintercept = 0.05, linetype = 2) +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5)) 

pl1
```


## plot 1 module time of day
```{r}
pl2 <- MEs.plot %>% 
  filter(str_detect(.$sampling_day, "[34]")) %>%
  ggplot(aes(x=sampling_time, y=blue, color=sampling_day)) +
  geom_smooth(aes(group=sampling_day)) +
  scale_color_viridis_d(name = "day") +
  ggtitle("blue", go.top6[go.top6$module=="blue", "top_6_go"]) +
  theme(plot.subtitle = element_text(size=9)) +
  ylab("expression") +
  xlab(NULL)

pl2
```
## plot 1 module by day
```{r}
pl3 <- MEs.plot %>%
  filter(sampling_time == "2_afternoon") %>%
  ggplot(aes(x=sampling_day, y=blue)) +
 # geom_boxplot() +
  geom_smooth(aes(group=sampling_time)) +
  ggtitle("blue") +
  ylab("expression") +
  xlab(NULL)

pl3
```

```{r}
(pl2 / pl3 | pl1) + plot_layout(widths = c(.8,.2))
```


## plotting function

```{r}
plot.me <- function(plot.module, plot.data, pvals){ 
  module.sym <- as.name(plot.module) # needed to programmatically select this using !!
  

  pl1 <- pvals %>% filter(module==plot.module) %>%
    ggplot(aes(y = fdr, x = term)) +
    geom_col(fill = plot.module) + 
    scale_y_continuous(transform=c("log10", "reverse")) +
    scale_x_discrete(limits = c("day", "time", "day:time")) +
    geom_hline(yintercept = 0.05, linetype = 2) +
    theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5)) 
  
  pl2 <-  plot.data %>% 
    filter(str_detect(.$sampling_day, "[34]")) %>%
    ggplot(aes(x=sampling_time, y=!!module.sym, color=sampling_day)) +
    geom_smooth(aes(group=sampling_day)) +
    scale_color_viridis_d(name = "day") +
    ggtitle(str_c(plot.module, ": time"), go.top6[go.top6$module==plot.module, "top_6_go"]) +
    theme(plot.subtitle = element_text(size=9)) +
    ylab("expression") +
    xlab(NULL)
  

  
  pl3 <- plot.data %>% 
  filter(sampling_time == "2_afternoon") %>%
  ggplot(aes(x=sampling_day, y=!!module.sym)) +
  geom_smooth(aes(group=sampling_time)) +
  ggtitle(str_c(plot.module, ": day")) +
  ylab("expression") +
  xlab(NULL)

  (pl2 / pl3 | pl1) + plot_layout(widths = c(.8,.2))

}

# test it

plot.me("black", MEs.plot, MEs.trt.aov)
```

```{r}
pdf(file="../output/WGCNA/ME_plots_live_genes_power5.pdf", height=8, width=12)



plotorder <- MEs.trt.aov %>% group_by(module) %>% slice_min(fdr, n=1, with_ties = FALSE) %>% arrange(fdr) %>% pull(module)
  
for(m in plotorder) {
  print(plot.me(m, MEs.plot, MEs.trt.aov))
}

dev.off()
```

