---
title: "20 Glucosinolate HeatMap"
output: html_notebook
---

Make a heatmap that show changes in expression of Dan's GSL gene list

```{r}
library(tidyverse)
library(gplots)
library(edgeR)
library(UpSetR)
```


## get gene IDs of GSL genes

```{r}
gsl.genes.agi <- read_csv("../input/GSL_genes_AGI.csv")
gsl.genes.agi
```

```{r}
annotation <- read_csv("../Annotation_copy/output/v3.0annotation/Brapa_V3.0_annotated.csv")
annotation
```

```{r}
gsl.genes <- gsl.genes.agi %>%
  left_join(annotation, by="AGI") %>% 
  filter(!is.na(name)) %>%
  select(AGI, GeneName, Pathways, transcript_ID=name, perc_ID, aln_length, eval) %>%
  group_by(transcript_ID) %>%
  slice_min(eval) %>% ungroup()
gsl.genes
```



```{r}
load("../output/TimeCourseDGE_LRT.Rdata")
```

```{r}
names(models)
DEgenes <- models %>% 
  mutate(toptags = map(lrt, ~ topTags(.,n=Inf)), #pull the DE gene info out of each model
         toptags = map(toptags, ~ .$table), # get the table of DE info
         toptags = map(toptags, ~ rownames_to_column(., "transcript_ID"))) %>%
  unnest(toptags) %>%
  group_by(transcript_ID) %>%
  filter(min(FDR) < 0.1) %>% # only keep transcripts where at least one test is < 0.1
  ungroup() %>%
  semi_join(gsl.genes, by="transcript_ID") %>%
  arrange(transcript_ID, FDR) %>% 
  select(day_time, transcript_ID, logFC, FDR)
DEgenes
```

## heat map of logFC

add gene metadata back in


```{r}
DEgenes <- DEgenes %>% left_join(gsl.genes) %>%
  mutate(label=str_c(transcript_ID,GeneName,Pathways, sep="_"),
         label=str_remove(label, "Bra"),
         label=str_remove(label, "\\.3C"))
DEgenes
```


make a matrix from DEgenes
```{r}
DEgenes.mat.trunc <- DEgenes %>%
  select(day_time, label, logFC) %>%
 # mutate(logFC=ifelse(abs(logFC) > 4, sign(logFC)*4, logFC)) %>%
  spread(key = day_time, value = logFC) %>%
  as.data.frame() %>%
  column_to_rownames("label") %>%
  as.matrix()
DEgenes.mat.trunc[1:10,1:10]
```

now, make the heat map

```{r, fig.height=8, fig.width=8}
DEgenes.mat.trunc %>% 
  heatmap.2(Colv = FALSE,
            keysize=1,
            dendrogram="row",
            sepwidth = c(0,0),
            margins=c(7,12),
            srtCol = 45,
            col=colorRampPalette(c("green", "black", "magenta")),
            trace = "none",
           # cexCol = 2
            #breaks=c(-8,-5, -2, -.5, .5, 2.5, 5, 10)
  )

```


# STOPPED HERE

heat map that clusters samples also:

```{r, fig.height=10, fig.width=10}
DEgenes.mat.trunc %>% 
  heatmap.2(keysize=1,
            dendrogram="both",
            sepwidth = c(0,0),
            labRow = FALSE,
            mar=c(12,0),
            srtCol = 45,
            col=colorRampPalette(c("green", "black", "magenta")),
            trace = "none",
            cexCol = 2
            #breaks=c(-8,-5, -2, -.5, .5, 2.5, 5, 10)
  )

```


count number of samples that each gene is DE in, and summarize

```{r}
DEgenes %>%
  summarize(n.timepoints=sum(FDR<0.05)) %>%
  group_by(n.timepoints) %>%
  summarize(n.genes=n()) %>%
  ggplot(aes(x=n.timepoints, y=n.genes)) +
  geom_col() +
  ggtitle("Number of genes detected at 1, 2, 3, ..., n time points") +
  ylab("number of genes") +
  xlab("number of time points")
```

Upset Plots
First calculate 0 / 1 table for DE at each time point

```{r}
DEgenes_01table_FDR0.05 <- DEgenes %>%
  ungroup() %>%
  mutate(DE=as.numeric(FDR<0.05)) %>%
  select(-logFC, -FDR) %>%
  spread(key = day_time, value = DE)  %>%
  mutate_at(.vars = vars(-transcript_ID), as.integer)
DEgenes_01table_FDR0.05
```

```{r, fig.width=8, fig.height=6}
DEgenes_01table_FDR0.05  %>% as.data.frame() %>% upset(nsets = 23, nintersects = 100,
                                                       order.by = "freq",
                                                       keep.order = TRUE, 
                                                       mb.ratio = c(.5,.5),
                                                       sets=colnames(DEgenes_01table_FDR0.05)[-1])
```


```{r}
DEgenes_01table_FDR0.1 <- models %>% 
  mutate(toptags = map(lrt, ~ topTags(.,n=Inf)), #pull the DE gene info out of each model
         toptags = map(toptags, ~ .$table), # get the table of DE info
         toptags = map(toptags, ~ rownames_to_column(., "transcript_ID"))) %>%
  unnest(toptags) %>%
  group_by(transcript_ID) %>%
  filter(min(FDR) < 0.1) %>% # only keep genes where FDR < 0.1 in at least one comparison.
  arrange(transcript_ID, FDR) %>% 
  select(day_time, transcript_ID, FDR) %>%
  ungroup() %>%
  mutate(DE=as.numeric(FDR<0.1)) %>%
  select(-FDR) %>%
  spread(key = day_time, value = DE) 
DEgenes_01table_FDR0.1
```

```{r, fig.width=8, fig.height=6}
DEgenes_01table_FDR0.1  %>% as.data.frame() %>% upset(nsets = 23, nintersects = 100,
                                                      order.by = "freq",
                                                      keep.order = TRUE, 
                                                      mb.ratio = c(.5,.5),
                                                      sets=colnames(DEgenes_01table_FDR0.1)[-1])
```

Look at overlap with exp13

```{r}
exp13 <- read_csv("../../Brapa_microbes/v3.0annotation/20180202-samples/output/DEG.e1.e3.complete.overlap.csv")
exp13
names(exp13)
```



```{r}
exp13_timecourse_.1 <- exp13 %>%
  select(transcript_ID, e1.root.trt5E_live.DEGs.add.rD.rFPsc.FDR0.1, e3.root.trtlive.DEGs.add.rD.rCR.v3.0anno.FDR0.1) %>%
  full_join(DEgenes_01table_FDR0.1) 
exp13_timecourse_.1[is.na(exp13_timecourse_.1)] <- 0
exp13_timecourse_.1 <- exp13_timecourse_.1 %>%
  mutate_at(.vars=vars(-transcript_ID), as.integer)
exp13_timecourse_.1
```


```{r, fig.width=12, fig.height=8}
exp13_timecourse_.1  %>% as.data.frame() %>% upset(nsets = 27,
                                                   nintersects = 85,
                                                   order.by = "freq",
                                                   group.by = "sets",
                                                   keep.order = TRUE, 
                                                   mb.ratio = c(.4,.6),
                                                   sets=colnames(exp13_timecourse_.1)[-1]
)
```


```{r}
exp13_timecourse_.1 %>% 
  filter(e1.root.trt5E_live.DEGs.add.rD.rFPsc.FDR0.1==1,
                               e3.root.trtlive.DEGs.add.rD.rCR.v3.0anno.FDR0.1==1,
                               X14.5_night_2==1)
```


What about upsetr plot for DE genes anywhere in time course?

```{r}
exp13 %>% select(transcript_ID, e1.root.trt5E_live.DEGs.add.rD.rFPsc.FDR0.1, e3.root.trtlive.DEGs.add.rD.rCR.v3.0anno.FDR0.1) %>%
  mutate(timecourse = as.numeric(transcript_ID %in% DEgenes_01table_FDR0.1$transcript_ID)) %>% 
  as.data.frame() %>%
  upset()
```

```{r}
exp13 %>% select(transcript_ID, e1.root.trt5E_live.DEGs.add.rD.rFPsc.FDR0.05, e3.root.trtlive.DEGs.add.rD.rCR.v3.0anno.FDR0.05) %>%
  mutate(timecourse = as.numeric(transcript_ID %in% DEgenes_01table_FDR0.05$transcript_ID)) %>% 
  as.data.frame() %>%
  upset()
```



```{r}
overlap %>%
  select(-total) %>%
  gather(key="set", value="genes", -day_time) %>%
  ggplot(aes(x=day_time, y=genes, fill=set)) +
  geom_col() +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust=0.5))
```

## Followup:

* histogram of FC, some concern about skewing
* upset plots
* heatmap with real expression values

FC histogram 
```{r}
DEgenes %>%
  ggplot(aes(x=logFC)) +
  geom_histogram(bins=50)
```

## heat map of cpm

```{r}
load("../output/timecourseDGE.Rdata")
```


```{r}
cpm <- cpmByGroup(dge)
cpm <- log2(cpm+1)
cpm <- cpm[unique(DEgenes$transcript_ID),]
colnames(cpm) <- colnames(cpm) %>% str_remove("FPsc-") %>%
  str_replace("(ATM_BLANK|SBC_OLD)-(.*)","\\2-\\1") %>%
  str_replace("ATM_BLANK", "BLNK") %>%
  str_replace("SBC_OLD", "SBC")
colorder <- c({colnames(cpm) %>% str_subset("BLNK") %>% sort()},
              {colnames(cpm) %>% str_subset("SBC") %>% sort()})
cpm <- cpm[,colorder]
dim(cpm)
```

```{r, fig.height=10, fig.width=10}
cpm %>% 
  heatmap.2(Colv = FALSE,
            scale = "row",
            keysize=1,
            dendrogram="row",
            sepwidth = c(0,0),
            labRow = FALSE,
            mar=c(12,0),
            srtCol = 45,
            col=colorRampPalette(c("green", "black", "magenta")),
            trace = "none",
            cexCol = 1,
            breaks=c(-8,-3:3, 8)
  )

```

Isn't really showing what I want.  The question is whether expression gets more similar because control becomes like treated or vice versa.

```{r}
cpm.long <- cpm %>% 
  as.data.frame() %>% 
  rownames_to_column(var="transcript_ID") %>% 
  gather(key="sample", value="log2cpm", -transcript_ID) %>%
  separate(sample, into=c("day","time","trt"), sep = "-", remove=FALSE) %>%
  mutate(day_time=str_c(day, "_", time))
cpm.long
```

take top 25 down and plot
```{r}
topdown <- DEgenes %>% filter(logFC<0) %>% arrange(FDR) %>% pull(transcript_ID) %>% unique() %>% magrittr::extract(1:25)

cpm.long %>%
  filter(transcript_ID %in% topdown,10) %>%
  ggplot(aes(x=day_time, y=log2cpm, group=trt, color=trt, lty=trt)) +
  geom_line() +
  facet_wrap(~transcript_ID) +
  ggtitle("Top down-regulated genes")
```

```{r}
topup <- DEgenes %>% filter(logFC>0) %>% arrange(FDR) %>% pull(transcript_ID) %>% unique() %>% magrittr::extract(1:25)

cpm.long %>%
  filter(transcript_ID %in% topup,10) %>%
  ggplot(aes(x=day_time, y=log2cpm, group=trt, color=trt, lty=trt)) +
  geom_line() +
  facet_wrap(~transcript_ID) +
  ggtitle("Top up-regulated genes")
```
